---
title: "canid_final"
author: "Dorothy Woo"
date: "2026-07-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## RSJZ0, UCL MRes Masters Dissertation (BIOS0014)
## Supervisors: Dr Lucas Porto, Dr Natalie Cooper (Natural History Museum, London)

# Load packages
```{r load packages, message=FALSE, warning=FALSE}
library(devtools)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(patchwork)
library(ggthemes)
library(lme4)
library("lmerTest")
library(emmeans)
library(vegan)
library(sf)
library(stringr)
library(terra)
library(purrr)
library(missMDA) 
library(FactoMineR) 
library(nnet) 
library(randomForest)
library(viridis)
library(factoextra)
library(knitr)
library(patchwork)
library(tidyr)
library(magick)
library(grid)
library(ggforce)
library(png)
library(Claddis)  
library(geoscale) 
library(dismo)  
library(ape)
library(tree)
library(funspace)
library(traitdata)
library(reticulate)
library(iucnsim)
library(rredlist)
library(corrplot)
library(psych)
library(scales)
library(plotly)
library(ggrepel)
library(geometry)
library(gridExtra)
library(svglite)
library(geiger)
library(cowplot)
library(tibble)
library(sp)
library(zoo)
library(gganimate)
library(phytools)
library(palaeoverse)
library(Rphylopars)
library(hypervolume)
library(rgl)
library(FNN)
library(picante)
library(lubridate)
```

# Data preparation and processing 
## Import data (data was manually compiled on Excel before imported to R)
```{r trait data processing, message=FALSE, warning=FALSE}
traits <- read_csv("selected_traits.csv") # please note that in this excel, I've also defined all extinction scenarios manually 
```

## Log transform all traits 
```{r transform data}
traits <- traits %>% 
  # log-transform
  mutate(across(
    .cols = c(
      RBL_original,
      Mass_original),
    .fns = ~log(.x), 
    .names = "{.col}_log"
  )) %>% 
  # rescale from 0 to 1
  mutate(across(
    .cols = ends_with("_log"),
    .fns = ~ scales::rescale(.x, to = c(0,1)),
    .names = "{.col}_scaled"
  )) 
```

## Prepare data frame for funspace
```{r funspace df}
# Compile a dataframe for funspace with relevant columns only
funspace_df <- traits %>% dplyr::select(
  Name, IUCN_status, continents, RBL_original_log_scaled, Mass_original_log_scaled,
  Extinction_CR_EN_NT, Extinction_SA_NT, Extinction_NA_NT, Extinction_Africa_NT, Extinction_Eurasia_NT) %>% 
  mutate(RBL = RBL_original_log_scaled, Mass = Mass_original_log_scaled) 
```

# Research Objective 1: How do extinction events affect FD globally
## 1.1: Global extinction of threatened and near-threatened species
```{r funspace global}
# Create a dataframe for funspace that only contains numeric values (RBL and Mass)
funspace_df_num <- funspace_df %>% dplyr::select(RBL, Mass)

# Construct trait space using funspace
traitspace <- funspace(x = funspace_df_num, n_divisions = 300)
summary(traitspace)
plot(x = traitspace, type = "global", quant.plot = TRUE, arrows = TRUE, arrows.length = 0.6, axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")

## Global extinction of CR EN NT species
traitspace_CR_EN_NT <-  funspace(x = funspace_df_num, group.vec = funspace_df$Extinction_CR_EN_NT, n_divisions = 300)
# Get summary statistics of the trait-space
summary(traitspace_CR_EN_NT)
# Plot trait-space density 
plot(x = traitspace_CR_EN_NT, type = "groups", quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")

# Compile results (FRic and FDiv)
global_fric_original <- traitspace$FD$global$FRich
global_fdiv_original <- traitspace$FD$global$FDiv
global_extinction_fric <- traitspace_CR_EN_NT$FD$groups$FRich["retained"]
global_extinction_fdiv <- traitspace_CR_EN_NT$FD$groups$FDiv["retained"]
global_fric_change <- 100*(global_extinction_fric-global_fric_original)/global_fric_original
globa_fdiv_change <- 100*(global_extinction_fdiv-global_fdiv_original)/global_fdiv_original

global_results <- tibble::tibble(
  Scenario = c("Global", "Removing CR, EN, NT"),
  FRic = c(global_fric_original, global_extinction_fric),
  FRic_change_percent = c("",global_fric_change),
  FDiv = c(global_fdiv_original, global_extinction_fdiv),
  FDiv_change_percent = c("", globa_fdiv_change)
)

global_results
```
## 1.2: Global trait-space: continent-specific extinction of threatened and near-threatened species
```{r funspace global continent specific}
# Construct trait space for each continent to visualise what regions of the trait-space does each continent occupy
traitspace_continent <-  funspace(x = funspace_df_num, PCs=c(1,2), group.vec = funspace_df$continents, n_divisions = 300)
plot(x = traitspace_continent, type = "groups", quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(traitspace_continent)

# Plot graphs in 2x2 format
par(mfrow=c(2,2))

# Removing threatened and near-threatened species in South America
traitspace_continent_SA <-  funspace(x = funspace_df_num, group.vec = funspace_df$Extinction_SA_NT, n_divisions = 300)
plot(x = traitspace_continent_SA, type = "groups", which.group = "retained", quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass", main = "South America")
title("South America only", cex.main = 1)
summary(traitspace_continent_SA)

# Removing threatened and near-threatened species in North America
traitspace_continent_NA <-  funspace(x = funspace_df_num, group.vec = funspace_df$Extinction_NA_NT, n_divisions = 300)
plot(x = traitspace_continent_NA, type = "groups", which.group = "retained", quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
title("North America only", cex.main = 1)
summary(traitspace_continent_NA)

# Removing threatened and near-threatened species in Africa
traitspace_continent_Africa <-  funspace(x = funspace_df_num, group.vec = funspace_df$Extinction_Africa_NT, n_divisions = 300)
plot(x = traitspace_continent_Africa, type = "groups", which.group = "retained", quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
title("Africa only", cex.main = 1)
summary(traitspace_continent_Africa)

# Removing threatened and near-threatened species in Eurasia
traitspace_continent_Eurasia <-  funspace(x = funspace_df_num, group.vec = funspace_df$Extinction_Eurasia_NT, n_divisions = 300)
plot(x = traitspace_continent_Eurasia, type = "groups", which.group = "retained", quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
title("Eurasia only", cex.main = 1)
summary(traitspace_continent_Eurasia)

# Compile results of FRic and FDiv
sa_fric <- traitspace_continent_SA$FD$groups$FRich["retained"]
sa_fdiv <- traitspace_continent_SA$FD$groups$FDiv["retained"]

na_fric <- traitspace_continent_NA$FD$groups$FRich["retained"]
na_fdiv <- traitspace_continent_NA$FD$groups$FDiv["retained"]

africa_fric <- traitspace_continent_Africa$FD$groups$FRich["retained"]
africa_fdiv <- traitspace_continent_Africa$FD$groups$FDiv["retained"]

eurasia_fric <- traitspace_continent_Eurasia$FD$groups$FRich["retained"]
eurasia_fdiv <- traitspace_continent_Eurasia$FD$groups$FDiv["retained"]

sa_fric_change <- 100*(sa_fric-global_fric_original)/global_fric_original
na_fric_change <- 100*(na_fric-global_fric_original)/global_fric_original
africa_fric_change <- 100*(africa_fric-global_fric_original)/global_fric_original
eurasia_fric_change <- 100*(eurasia_fric-global_fric_original)/global_fric_original

sa_fdiv_change <- 100*(sa_fdiv-global_fdiv_original)/global_fdiv_original
na_fdiv_change <- 100*(na_fdiv-global_fdiv_original)/global_fdiv_original
africa_fdiv_change <- 100*(africa_fdiv-global_fdiv_original)/global_fdiv_original
eurasia_fdiv_change <- 100*(eurasia_fdiv-global_fdiv_original)/global_fdiv_original

continental_results <- tibble::tibble(
  Continent = c("Global", "South America", "North America",  "Africa", "Eurasia"),
  FRic = c(global_fric_original, sa_fric, na_fric, africa_fric, eurasia_fric),
  FRic_change_percent = c("",sa_fric_change, na_fric_change, 
                          africa_fric_change,eurasia_fric_change),
  FDiv = c(global_fdiv_original, sa_fdiv, na_fdiv, africa_fdiv, eurasia_fdiv),
  FDiv_change_percent = c("",sa_fdiv_change, na_fdiv_change, 
                          africa_fdiv_change,eurasia_fdiv_change),
)

continental_results
```

# Research Objective 2: How do extinction events affect FD continentally 
```{r continent separate results}
# Plot graphs in 2x2 format
par(mfrow = c(2,2))

# South America trait-space and extinction
df_temp <- funspace_df %>% filter(continents == "South America") # filter only species found in South America to construct relevant trait-space
df_temp_num <- df_temp %>% dplyr::select(RBL, Mass)
temp_funspace <- funspace(x = df_temp_num, group.vec = df_temp$Extinction_SA_NT, n_divisions = 300)
summary(temp_funspace)
# Extract relevant statistics from the trait-space
SA_cfrich_original <- temp_funspace$FD$global$FRich
SA_cfrich_new <- temp_funspace$FD$groups$FRich
SA_cfdiv_original <- temp_funspace$FD$global$FDiv
SA_cfdiv_new <- temp_funspace$FD$groups$FDiv
# Plot trait-space density for the continent
plot(x = temp_funspace, type = "groups", which.group = "retained", quant.plot = TRUE, globalContour = T, colors = c("cadetblue1", "blue"), pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass") 
title("South America", cex.main = 1) # Add title to the graph

# North America trait-space and extinction
df_temp <- funspace_df %>% filter(continents == "North America") 
df_temp_num <- df_temp %>% dplyr::select(RBL, Mass)
temp_funspace <- funspace(x = df_temp_num, group.vec = df_temp$Extinction_NA_NT, n_divisions = 300)
summary(temp_funspace)
NA_cfrich_original <- temp_funspace$FD$global$FRich
NA_cfrich_new <- temp_funspace$FD$groups$FRich
NA_cfdiv_original <- temp_funspace$FD$global$FDiv
NA_cfdiv_new <- temp_funspace$FD$groups$FDiv
plot(x = temp_funspace, type = "groups", which.group = "retained", quant.plot = TRUE, globalContour = T, colors = c("cadetblue1", "blue"), pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
title("North America", cex.main = 1)


# Africa trait-space and extinction
df_temp <- funspace_df %>% filter(continents == "Africa")
df_temp_num <- df_temp %>% dplyr::select(RBL, Mass)
temp_funspace <- funspace(x = df_temp_num, group.vec = df_temp$Extinction_Africa_NT, n_divisions = 300)
summary(temp_funspace)
Africa_cfrich_original <- temp_funspace$FD$global$FRich
Africa_cfrich_new <- temp_funspace$FD$groups$FRich
Africa_cfdiv_original <- temp_funspace$FD$global$FDiv
Africa_cfdiv_new <- temp_funspace$FD$groups$FDiv
plot(x = temp_funspace, type = "groups", which.group = "retained", quant.plot = TRUE, globalContour = T, colors = c("cadetblue1", "blue"), pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
title("Africa", cex.main = 1)

# Eurasia trait-space and extinction
df_temp <- funspace_df %>% filter(continents == "Eurasia")
df_temp_num <- df_temp %>% dplyr::select(RBL, Mass)
temp_funspace <- funspace(x = df_temp_num, group.vec = df_temp$Extinction_Eurasia_NT, n_divisions = 300)
summary(temp_funspace)
Eurasia_cfrich_original <- temp_funspace$FD$global$FRich
Eurasia_cfrich_new <- temp_funspace$FD$groups$FRich
Eurasia_cfdiv_original <- temp_funspace$FD$global$FDiv
Eurasia_cfdiv_new <- temp_funspace$FD$groups$FDiv
plot(x = temp_funspace, type = "groups", which.group = "retained", quant.plot = TRUE, globalContour = T, colors = c("cadetblue1", "blue"), pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
title("Eurasia", cex.main = 1)

# FRic and FDiv percentage change results
SA_cfrich_new <- as.numeric(SA_cfrich_new["retained"])
SA_cfdiv_new <- as.numeric(SA_cfdiv_new["retained"])
SA_cfrich_percent <- ((SA_cfrich_new-SA_cfrich_original)/SA_cfrich_original)*100
SA_cfdiv_percent <- ((SA_cfdiv_new-SA_cfdiv_original)/SA_cfdiv_original)*100

NA_cfrich_new <- as.numeric(NA_cfrich_new["retained"])
NA_cfdiv_new <- as.numeric(NA_cfdiv_new["retained"])
NA_cfrich_percent <- ((NA_cfrich_new-NA_cfrich_original)/NA_cfrich_original)*100
NA_cfdiv_percent <- ((NA_cfdiv_new-NA_cfdiv_original)/NA_cfdiv_original)*100

Africa_cfrich_new <- as.numeric(Africa_cfrich_new["retained"])
Africa_cfdiv_new <- as.numeric(Africa_cfdiv_new["retained"])
Africa_cfrich_percent <- ((Africa_cfrich_new-Africa_cfrich_original)/Africa_cfrich_original)*100
Africa_cfdiv_percent <- ((Africa_cfdiv_new-Africa_cfdiv_original)/Africa_cfdiv_original)*100

Eurasia_cfrich_new <- as.numeric(Eurasia_cfrich_new["retained"])
Eurasia_cfdiv_new <- as.numeric(Eurasia_cfdiv_new["retained"])
Eurasia_cfrich_percent <- ((Eurasia_cfrich_new-Eurasia_cfrich_original)/Eurasia_cfrich_original)*100
Eurasia_cfdiv_percent <- ((Eurasia_cfdiv_new-Eurasia_cfdiv_original)/Eurasia_cfdiv_original)*100

# Compile results
continent_standalone_results <- tibble::tibble(
  Continent = c("South America", "South America", "North America", "North America", "Africa","Africa", "Eurasia", "Eurasia"),
  Scenario = c("Original", "Post-extinction", "Original", "Post-extinction", "Original", "Post-extinction", "Original", "Post-extinction"),
  FRic = c(SA_cfrich_original, SA_cfrich_new, NA_cfrich_original, NA_cfrich_new, 
           Africa_cfrich_original, Africa_cfrich_new, Eurasia_cfrich_original, Eurasia_cfrich_new),
  Change_in_FRic = c("-", SA_cfrich_percent, "-", NA_cfrich_percent, "-", Africa_cfrich_percent, "-", Eurasia_cfrich_percent),
  FDiv = c(SA_cfdiv_original, SA_cfdiv_new, NA_cfdiv_original, NA_cfdiv_new, 
           Africa_cfdiv_original, Africa_cfdiv_new, Eurasia_cfdiv_original, Eurasia_cfdiv_new),
  Change_in_FDiv = c("-", SA_cfdiv_percent, "-", NA_cfdiv_percent, "-", Africa_cfdiv_percent, "-", Eurasia_cfdiv_percent))

continent_standalone_results
```

# Sensitivity check for Research Objectives 1 and 2 using probability thresholds of 0.99 and 0.995
## Using probability threshold of 0.99
```{r sensitivity check 0.99}
### Part 1: Global results ###
## Part 1.1: Global extinction of CR, EN, NT ##
# Construct a traitspace with probability threshold 0.99
traitspace_0.99 <- funspace(x = funspace_df_num, n_divisions = 300, threshold=0.99)
summary(traitspace_0.99)
# Construct a traitspace with probability threshold 0.99 after removal of CR, EN, NT species
traitspace_CR_EN_NT_0.99 <-  funspace(x = funspace_df_num, group.vec = funspace_df$Extinction_CR_EN_NT, n_divisions = 300, threshold=0.99)
summary(traitspace_CR_EN_NT_0.99)
# Collate all stats from the 2 trait spaces above
global_fric_original_0.99 <- traitspace_0.99$FD$global$FRich
global_fdiv_original_0.99 <- traitspace_0.99$FD$global$FDiv
global_extinction_fric_0.99 <- traitspace_CR_EN_NT_0.99$FD$groups$FRich["retained"]
global_extinction_fdiv_0.99 <- traitspace_CR_EN_NT_0.99$FD$groups$FDiv["retained"]
global_fric_change_0.99 <- 100*(global_extinction_fric_0.99-global_fric_original_0.99)/global_fric_original_0.99
globa_fdiv_change_0.99 <- 100*(global_extinction_fdiv_0.99-global_fdiv_original_0.99)/global_fdiv_original_0.99

global_results_0.99 <- tibble::tibble(
  Scenario = c("Global", "Removing CR, EN, NT"),
  FRic = c(global_fric_original_0.99, global_extinction_fric_0.99),
  FRic_change_percent = c("",global_fric_change_0.99),
  FDiv = c(global_fdiv_original_0.99, global_extinction_fdiv_0.99),
  FDiv_change_percent = c("", globa_fdiv_change_0.99)
)


## Part 1.2: Global trait space of continental extinction ##
traitspace_continent_0.99 <-  funspace(x = funspace_df_num, PCs=c(1,2), group.vec = funspace_df$continents, n_divisions = 300, threshold=0.99)
summary(traitspace_continent_0.99)

# Removing threatened and near-threatened species in South America
traitspace_continent_SA_0.99 <-  funspace(x = funspace_df_num, group.vec = funspace_df$Extinction_SA_NT, n_divisions = 300,threshold=0.99)
summary(traitspace_continent_SA_0.99)

# Removing threatened and near-threatened species in North America
traitspace_continent_NA_0.99 <-  funspace(x = funspace_df_num, group.vec = funspace_df$Extinction_NA_NT, n_divisions = 300,threshold=0.99)
summary(traitspace_continent_NA_0.99)

# Removing threatened and near-threatened species in Africa
traitspace_continent_Africa_0.99 <-  funspace(x = funspace_df_num, group.vec = funspace_df$Extinction_Africa_NT, n_divisions = 300,threshold=0.99)
summary(traitspace_continent_Africa_0.99)

# Removing threatened and near-threatened species in Eurasia
traitspace_continent_Eurasia_0.99 <-  funspace(x = funspace_df_num, group.vec = funspace_df$Extinction_Eurasia_NT, n_divisions = 300,threshold=0.99)
summary(traitspace_continent_Eurasia_0.99)

# Compile results of FRic and FDiv
sa_fric_0.99 <- traitspace_continent_SA_0.99$FD$groups$FRich["retained"]
sa_fdiv_0.99 <- traitspace_continent_SA_0.99$FD$groups$FDiv["retained"]

na_fric_0.99 <- traitspace_continent_NA_0.99$FD$groups$FRich["retained"]
na_fdiv_0.99 <- traitspace_continent_NA_0.99$FD$groups$FDiv["retained"]

africa_fric_0.99 <- traitspace_continent_Africa_0.99$FD$groups$FRich["retained"]
africa_fdiv_0.99 <- traitspace_continent_Africa_0.99$FD$groups$FDiv["retained"]

eurasia_fric_0.99 <- traitspace_continent_Eurasia_0.99$FD$groups$FRich["retained"]
eurasia_fdiv_0.99 <- traitspace_continent_Eurasia_0.99$FD$groups$FDiv["retained"]

sa_fric_change_0.99 <- 100*(sa_fric_0.99-global_fric_original_0.99)/global_fric_original_0.99
na_fric_change_0.99 <- 100*(na_fric_0.99-global_fric_original_0.99)/global_fric_original_0.99
africa_fric_change_0.99 <- 100*(africa_fric_0.99-global_fric_original_0.99)/global_fric_original_0.99
eurasia_fric_change_0.99 <- 100*(eurasia_fric_0.99-global_fric_original_0.99)/global_fric_original_0.99

sa_fdiv_change_0.99 <- 100*(sa_fdiv_0.99-global_fdiv_original_0.99)/global_fdiv_original_0.99
na_fdiv_change_0.99 <- 100*(na_fdiv_0.99-global_fdiv_original_0.99)/global_fdiv_original_0.99
africa_fdiv_change_0.99 <- 100*(africa_fdiv_0.99-global_fdiv_original_0.99)/global_fdiv_original_0.99
eurasia_fdiv_change_0.99 <- 100*(eurasia_fdiv_0.99-global_fdiv_original_0.99)/global_fdiv_original_0.99

continental_results_0.99 <- tibble::tibble(
  Continent = c("Global", "South America", "North America",  "Africa", "Eurasia"),
  FRic = c(global_fric_original_0.99, sa_fric_0.99, na_fric_0.99, africa_fric_0.99, eurasia_fric_0.99),
  FRic_change_percent = c("",sa_fric_change_0.99, na_fric_change_0.99, 
                          africa_fric_change_0.99,eurasia_fric_change_0.99),
  FDiv = c(global_fdiv_original_0.99, sa_fdiv_0.99, na_fdiv_0.99, africa_fdiv_0.99, eurasia_fdiv_0.99),
  FDiv_change_percent = c("",sa_fdiv_change_0.99, na_fdiv_change_0.99, 
                          africa_fdiv_change_0.99,eurasia_fdiv_change_0.99),
)

continental_results_0.99

### Part 2: Standalone continental results ### 

# South America trait-space and extinction
df_temp <- funspace_df %>% filter(continents == "South America") # filter only species found in South America to construct relevant trait-space
df_temp_num <- df_temp %>% dplyr::select(RBL, Mass)
temp_funspace_0.99 <- funspace(x = df_temp_num, group.vec = df_temp$Extinction_SA_NT, n_divisions = 300, threshold=0.99)
summary(temp_funspace_0.99)
# Extract relevant statistics from the trait-space
SA_cfrich_original_0.99 <- temp_funspace_0.99$FD$global$FRich
SA_cfrich_new_0.99 <- temp_funspace_0.99$FD$groups$FRich
SA_cfdiv_original_0.99 <- temp_funspace_0.99$FD$global$FDiv
SA_cfdiv_new_0.99 <- temp_funspace_0.99$FD$groups$FDiv

# North America trait-space and extinction
df_temp <- funspace_df %>% filter(continents == "North America") 
df_temp_num <- df_temp %>% dplyr::select(RBL, Mass)
temp_funspace_0.99 <- funspace(x = df_temp_num, group.vec = df_temp$Extinction_NA_NT, n_divisions = 300, threshold=0.99)
summary(temp_funspace_0.99)
NA_cfrich_original_0.99 <- temp_funspace_0.99$FD$global$FRich
NA_cfrich_new_0.99 <- temp_funspace_0.99$FD$groups$FRich
NA_cfdiv_original_0.99 <- temp_funspace_0.99$FD$global$FDiv
NA_cfdiv_new_0.99 <- temp_funspace_0.99$FD$groups$FDiv

# Africa trait-space and extinction
df_temp <- funspace_df %>% filter(continents == "Africa")
df_temp_num <- df_temp %>% dplyr::select(RBL, Mass)
temp_funspace_0.99 <- funspace(x = df_temp_num, group.vec = df_temp$Extinction_Africa_NT, n_divisions = 300, threshold=0.99)
summary(temp_funspace_0.99)
Africa_cfrich_original_0.99 <- temp_funspace_0.99$FD$global$FRich
Africa_cfrich_new_0.99 <- temp_funspace_0.99$FD$groups$FRich
Africa_cfdiv_original_0.99 <- temp_funspace_0.99$FD$global$FDiv
Africa_cfdiv_new_0.99 <- temp_funspace_0.99$FD$groups$FDiv

# Eurasia trait-space and extinction
df_temp <- funspace_df %>% filter(continents == "Eurasia")
df_temp_num <- df_temp %>% dplyr::select(RBL, Mass)
temp_funspace_0.99 <- funspace(x = df_temp_num, group.vec = df_temp$Extinction_Eurasia_NT, n_divisions = 300, threshold=0.99)
summary(temp_funspace_0.99)
Eurasia_cfrich_original_0.99 <- temp_funspace_0.99$FD$global$FRich
Eurasia_cfrich_new_0.99 <- temp_funspace_0.99$FD$groups$FRich
Eurasia_cfdiv_original_0.99 <- temp_funspace_0.99$FD$global$FDiv
Eurasia_cfdiv_new_0.99 <- temp_funspace_0.99$FD$groups$FDiv

SA_cfrich_new_0.99 <- as.numeric(SA_cfrich_new_0.99["retained"])
SA_cfdiv_new_0.99 <- as.numeric(SA_cfdiv_new_0.99["retained"])
SA_cfrich_percent_0.99 <- ((SA_cfrich_new_0.99-SA_cfrich_original_0.99)/SA_cfrich_original_0.99)*100
SA_cfdiv_percent_0.99 <- ((SA_cfdiv_new_0.99-SA_cfdiv_original_0.99)/SA_cfdiv_original_0.99)*100

NA_cfrich_new_0.99 <- as.numeric(NA_cfrich_new_0.99["retained"])
NA_cfdiv_new_0.99 <- as.numeric(NA_cfdiv_new_0.99["retained"])
NA_cfrich_percent_0.99 <- ((NA_cfrich_new_0.99-NA_cfrich_original_0.99)/NA_cfrich_original_0.99)*100
NA_cfdiv_percent_0.99 <- ((NA_cfdiv_new_0.99-NA_cfdiv_original_0.99)/NA_cfdiv_original_0.99)*100

Africa_cfrich_new_0.99 <- as.numeric(Africa_cfrich_new_0.99["retained"])
Africa_cfdiv_new_0.99 <- as.numeric(Africa_cfdiv_new_0.99["retained"])
Africa_cfrich_percent_0.99 <- ((Africa_cfrich_new_0.99-Africa_cfrich_original_0.99)/Africa_cfrich_original_0.99)*100
Africa_cfdiv_percent_0.99 <- ((Africa_cfdiv_new_0.99-Africa_cfdiv_original_0.99)/Africa_cfdiv_original_0.99)*100

Eurasia_cfrich_new_0.99 <- as.numeric(Eurasia_cfrich_new_0.99["retained"])
Eurasia_cfdiv_new_0.99 <- as.numeric(Eurasia_cfdiv_new_0.99["retained"])
Eurasia_cfrich_percent_0.99 <- ((Eurasia_cfrich_new_0.99-Eurasia_cfrich_original_0.99)/Eurasia_cfrich_original_0.99)*100
Eurasia_cfdiv_percent_0.99 <- ((Eurasia_cfdiv_new_0.99-Eurasia_cfdiv_original_0.99)/Eurasia_cfdiv_original_0.99)*100

continent_standalone_results_0.99 <- tibble::tibble(
  Continent = c("South America", "South America", "North America", "North America", "Africa","Africa", "Eurasia", "Eurasia"),
  Scenario = c("Original", "Post-extinction", "Original", "Post-extinction", "Original", "Post-extinction", "Original", "Post-extinction"),
  FRic = c(SA_cfrich_original_0.99, SA_cfrich_new_0.99, NA_cfrich_original_0.99, NA_cfrich_new_0.99, 
           Africa_cfrich_original_0.99, Africa_cfrich_new_0.99, Eurasia_cfrich_original_0.99, Eurasia_cfrich_new_0.99),
  Change_in_FRic = c("-", SA_cfrich_percent_0.99, "-", NA_cfrich_percent_0.99, "-", Africa_cfrich_percent_0.99, "-", Eurasia_cfrich_percent_0.99),
  FDiv = c(SA_cfdiv_original_0.99, SA_cfdiv_new_0.99, NA_cfdiv_original_0.99, NA_cfdiv_new_0.99, 
           Africa_cfdiv_original_0.99, Africa_cfdiv_new_0.99, Eurasia_cfdiv_original_0.99, Eurasia_cfdiv_new_0.99),
  Change_in_FDiv = c("-", SA_cfdiv_percent_0.99, "-", NA_cfdiv_percent_0.99, "-", Africa_cfdiv_percent_0.99, "-", Eurasia_cfdiv_percent_0.99))

### All results from Parts 1 and 2 ### 
global_results_0.99
continental_results_0.99
continent_standalone_results_0.99
```

2. Using probability threshold of 0.995
```{r sensitivity check 0.995}
### Part 1: Global Results ###
## Part 1.1: Global extinction of CR, EN, NT ##
# Construct a traitspace with probability threshold 0.995
traitspace_0.995 <- funspace(x = funspace_df_num, n_divisions = 300, threshold=0.995)
summary(traitspace_0.995)
# Construct a traitspace with probability threshold 0.995 after removal of CR, EN, NT species
traitspace_CR_EN_NT_0.995 <-  funspace(x = funspace_df_num, group.vec = funspace_df$Extinction_CR_EN_NT, n_divisions = 300, threshold=0.995)
summary(traitspace_CR_EN_NT_0.995)
# Collate all stats from the 2 trait spaces above
global_fric_original_0.995 <- traitspace_0.995$FD$global$FRich
global_fdiv_original_0.995 <- traitspace_0.995$FD$global$FDiv
global_extinction_fric_0.995 <- traitspace_CR_EN_NT_0.995$FD$groups$FRich["retained"]
global_extinction_fdiv_0.995 <- traitspace_CR_EN_NT_0.995$FD$groups$FDiv["retained"]
global_fric_change_0.995 <- 100*(global_extinction_fric_0.995-global_fric_original_0.995)/global_fric_original_0.995
globa_fdiv_change_0.995 <- 100*(global_extinction_fdiv_0.995-global_fdiv_original_0.995)/global_fdiv_original_0.995

global_results_0.995 <- tibble::tibble(
  Scenario = c("Global", "Removing CR, EN, NT"),
  FRic = c(global_fric_original_0.995, global_extinction_fric_0.995),
  FRic_change_percent = c("",global_fric_change_0.995),
  FDiv = c(global_fdiv_original_0.995, global_extinction_fdiv_0.995),
  FDiv_change_percent = c("", globa_fdiv_change_0.995)
)


## Part 1.2: Global trait space of continental extinction ##
traitspace_continent_0.995 <-  funspace(x = funspace_df_num, PCs=c(1,2), group.vec = funspace_df$continents, n_divisions = 300, threshold=0.995)
summary(traitspace_continent_0.995)

# Removing threatened and near-threatened species in South America
traitspace_continent_SA_0.995 <-  funspace(x = funspace_df_num, group.vec = funspace_df$Extinction_SA_NT, n_divisions = 300,threshold=0.995)
summary(traitspace_continent_SA_0.995)

# Removing threatened and near-threatened species in North America
traitspace_continent_NA_0.995 <-  funspace(x = funspace_df_num, group.vec = funspace_df$Extinction_NA_NT, n_divisions = 300,threshold=0.995)
summary(traitspace_continent_NA_0.995)

# Removing threatened and near-threatened species in Africa
traitspace_continent_Africa_0.995 <-  funspace(x = funspace_df_num, group.vec = funspace_df$Extinction_Africa_NT, n_divisions = 300,threshold=0.995)
summary(traitspace_continent_Africa_0.995)

# Removing threatened and near-threatened species in Eurasia
traitspace_continent_Eurasia_0.995 <-  funspace(x = funspace_df_num, group.vec = funspace_df$Extinction_Eurasia_NT, n_divisions = 300,threshold=0.995)
summary(traitspace_continent_Eurasia_0.995)

# Compile results of FRic and FDiv
sa_fric_0.995 <- traitspace_continent_SA_0.995$FD$groups$FRich["retained"]
sa_fdiv_0.995 <- traitspace_continent_SA_0.995$FD$groups$FDiv["retained"]

na_fric_0.995 <- traitspace_continent_NA_0.995$FD$groups$FRich["retained"]
na_fdiv_0.995 <- traitspace_continent_NA_0.995$FD$groups$FDiv["retained"]

africa_fric_0.995 <- traitspace_continent_Africa_0.995$FD$groups$FRich["retained"]
africa_fdiv_0.995 <- traitspace_continent_Africa_0.995$FD$groups$FDiv["retained"]

eurasia_fric_0.995 <- traitspace_continent_Eurasia_0.995$FD$groups$FRich["retained"]
eurasia_fdiv_0.995 <- traitspace_continent_Eurasia_0.995$FD$groups$FDiv["retained"]

sa_fric_change_0.995 <- 100*(sa_fric_0.995-global_fric_original_0.995)/global_fric_original_0.995
na_fric_change_0.995 <- 100*(na_fric_0.995-global_fric_original_0.995)/global_fric_original_0.995
africa_fric_change_0.995 <- 100*(africa_fric_0.995-global_fric_original_0.995)/global_fric_original_0.995
eurasia_fric_change_0.995 <- 100*(eurasia_fric_0.995-global_fric_original_0.995)/global_fric_original_0.995

sa_fdiv_change_0.995 <- 100*(sa_fdiv_0.995-global_fdiv_original_0.995)/global_fdiv_original_0.995
na_fdiv_change_0.995 <- 100*(na_fdiv_0.995-global_fdiv_original_0.995)/global_fdiv_original_0.995
africa_fdiv_change_0.995 <- 100*(africa_fdiv_0.995-global_fdiv_original_0.995)/global_fdiv_original_0.995
eurasia_fdiv_change_0.995 <- 100*(eurasia_fdiv_0.995-global_fdiv_original_0.995)/global_fdiv_original_0.995

continental_results_0.995 <- tibble::tibble(
  Continent = c("Global", "South America", "North America",  "Africa", "Eurasia"),
  FRic = c(global_fric_original_0.995, sa_fric_0.995, na_fric_0.995, africa_fric_0.995, eurasia_fric_0.995),
  FRic_change_percent = c("",sa_fric_change_0.995, na_fric_change_0.995, 
                          africa_fric_change_0.995,eurasia_fric_change_0.995),
  FDiv = c(global_fdiv_original_0.995, sa_fdiv_0.995, na_fdiv_0.995, africa_fdiv_0.995, eurasia_fdiv_0.995),
  FDiv_change_percent = c("",sa_fdiv_change_0.995, na_fdiv_change_0.995, 
                          africa_fdiv_change_0.995,eurasia_fdiv_change_0.995),
)

continental_results_0.995

### Part 2: Standalone continental results ### 

# South America trait-space and extinction
df_temp <- funspace_df %>% filter(continents == "South America") # filter only species found in South America to construct relevant trait-space
df_temp_num <- df_temp %>% dplyr::select(RBL, Mass)
temp_funspace_0.995 <- funspace(x = df_temp_num, group.vec = df_temp$Extinction_SA_NT, n_divisions = 300, threshold=0.995)
summary(temp_funspace_0.995)
# Extract relevant statistics from the trait-space
SA_cfrich_original_0.995 <- temp_funspace_0.995$FD$global$FRich
SA_cfrich_new_0.995 <- temp_funspace_0.995$FD$groups$FRich
SA_cfdiv_original_0.995 <- temp_funspace_0.995$FD$global$FDiv
SA_cfdiv_new_0.995 <- temp_funspace_0.995$FD$groups$FDiv

# North America trait-space and extinction
df_temp <- funspace_df %>% filter(continents == "North America") 
df_temp_num <- df_temp %>% dplyr::select(RBL, Mass)
temp_funspace_0.995 <- funspace(x = df_temp_num, group.vec = df_temp$Extinction_NA_NT, n_divisions = 300, threshold=0.995)
summary(temp_funspace_0.995)
NA_cfrich_original_0.995 <- temp_funspace_0.995$FD$global$FRich
NA_cfrich_new_0.995 <- temp_funspace_0.995$FD$groups$FRich
NA_cfdiv_original_0.995 <- temp_funspace_0.995$FD$global$FDiv
NA_cfdiv_new_0.995 <- temp_funspace_0.995$FD$groups$FDiv

# Africa trait-space and extinction
df_temp <- funspace_df %>% filter(continents == "Africa")
df_temp_num <- df_temp %>% dplyr::select(RBL, Mass)
temp_funspace_0.995 <- funspace(x = df_temp_num, group.vec = df_temp$Extinction_Africa_NT, n_divisions = 300, threshold=0.995)
summary(temp_funspace_0.995)
Africa_cfrich_original_0.995 <- temp_funspace_0.995$FD$global$FRich
Africa_cfrich_new_0.995 <- temp_funspace_0.995$FD$groups$FRich
Africa_cfdiv_original_0.995 <- temp_funspace_0.995$FD$global$FDiv
Africa_cfdiv_new_0.995 <- temp_funspace_0.995$FD$groups$FDiv

# Eurasia trait-space and extinction
df_temp <- funspace_df %>% filter(continents == "Eurasia")
df_temp_num <- df_temp %>% dplyr::select(RBL, Mass)
temp_funspace_0.995 <- funspace(x = df_temp_num, group.vec = df_temp$Extinction_Eurasia_NT, n_divisions = 300, threshold=0.995)
summary(temp_funspace_0.995)
Eurasia_cfrich_original_0.995 <- temp_funspace_0.995$FD$global$FRich
Eurasia_cfrich_new_0.995 <- temp_funspace_0.995$FD$groups$FRich
Eurasia_cfdiv_original_0.995 <- temp_funspace_0.995$FD$global$FDiv
Eurasia_cfdiv_new_0.995 <- temp_funspace_0.995$FD$groups$FDiv

SA_cfrich_new_0.995 <- as.numeric(SA_cfrich_new_0.995["retained"])
SA_cfdiv_new_0.995 <- as.numeric(SA_cfdiv_new_0.995["retained"])
SA_cfrich_percent_0.995 <- ((SA_cfrich_new_0.995-SA_cfrich_original_0.995)/SA_cfrich_original_0.995)*100
SA_cfdiv_percent_0.995 <- ((SA_cfdiv_new_0.995-SA_cfdiv_original_0.995)/SA_cfdiv_original_0.995)*100

NA_cfrich_new_0.995 <- as.numeric(NA_cfrich_new_0.995["retained"])
NA_cfdiv_new_0.995 <- as.numeric(NA_cfdiv_new_0.995["retained"])
NA_cfrich_percent_0.995 <- ((NA_cfrich_new_0.995-NA_cfrich_original_0.995)/NA_cfrich_original_0.995)*100
NA_cfdiv_percent_0.995 <- ((NA_cfdiv_new_0.995-NA_cfdiv_original_0.995)/NA_cfdiv_original_0.995)*100

Africa_cfrich_new_0.995 <- as.numeric(Africa_cfrich_new_0.995["retained"])
Africa_cfdiv_new_0.995 <- as.numeric(Africa_cfdiv_new_0.995["retained"])
Africa_cfrich_percent_0.995 <- ((Africa_cfrich_new_0.995-Africa_cfrich_original_0.995)/Africa_cfrich_original_0.995)*100
Africa_cfdiv_percent_0.995 <- ((Africa_cfdiv_new_0.995-Africa_cfdiv_original_0.995)/Africa_cfdiv_original_0.995)*100

Eurasia_cfrich_new_0.995 <- as.numeric(Eurasia_cfrich_new_0.995["retained"])
Eurasia_cfdiv_new_0.995 <- as.numeric(Eurasia_cfdiv_new_0.995["retained"])
Eurasia_cfrich_percent_0.995 <- ((Eurasia_cfrich_new_0.995-Eurasia_cfrich_original_0.995)/Eurasia_cfrich_original_0.995)*100
Eurasia_cfdiv_percent_0.995 <- ((Eurasia_cfdiv_new_0.995-Eurasia_cfdiv_original_0.995)/Eurasia_cfdiv_original_0.995)*100

continent_standalone_results_0.995 <- tibble::tibble(
  Continent = c("South America", "South America", "North America", "North America", "Africa","Africa", "Eurasia", "Eurasia"),
  Scenario = c("Original", "Post-extinction", "Original", "Post-extinction", "Original", "Post-extinction", "Original", "Post-extinction"),
  FRic = c(SA_cfrich_original_0.995, SA_cfrich_new_0.995, NA_cfrich_original_0.995, NA_cfrich_new_0.995, 
           Africa_cfrich_original_0.995, Africa_cfrich_new_0.995, Eurasia_cfrich_original_0.995, Eurasia_cfrich_new_0.995),
  Change_in_FRic = c("-", SA_cfrich_percent_0.995, "-", NA_cfrich_percent_0.995, "-", Africa_cfrich_percent_0.995, "-", Eurasia_cfrich_percent_0.995),
  FDiv = c(SA_cfdiv_original_0.995, SA_cfdiv_new_0.995, NA_cfdiv_original_0.995, NA_cfdiv_new_0.995, 
           Africa_cfdiv_original_0.995, Africa_cfdiv_new_0.995, Eurasia_cfdiv_original_0.995, Eurasia_cfdiv_new_0.995),
  Change_in_FDiv = c("-", SA_cfdiv_percent_0.995, "-", NA_cfdiv_percent_0.995, "-", Africa_cfdiv_percent_0.995, "-", Eurasia_cfdiv_percent_0.995))

### All results from Parts 1 and 2 ### 
global_results_0.995
continental_results_0.995
continent_standalone_results_0.995
```

# Supplementary analyses: How do contemporary, anthropogenic extinction events compare to past historical events?
## Use PBDB database to get continent and time information
```{r pbdb, warning=FALSE, message=FALSE}
# Import PBDB information
pbdb <- read_csv("pbdb_data.csv")
pbdb <- pbdb %>% dplyr::select(Name, lng, lat, max_ma, min_ma) %>% mutate(mid_ma = (max_ma+min_ma)/2)

# Convert longitude and latitude points to sf points 
pbdb_sfpts <- st_as_sf(pbdb, coords = c("lng","lat"), crs = 4326)

# Assign a period for each species
bins <- time_bins(scale = "international ages")
pbdb <- bin_time(
  occdf = pbdb,
  bins = bins,
  min_ma = "min_ma",
  max_ma = "max_ma",
  method = "majority"
)
bins_name <- dplyr::select(bins, bin, interval_name) 
pbdb <- left_join(pbdb, bins, by=c("bin_assignment" = "bin"))
pbdb <- pbdb %>% mutate(period = interval_name) 
pbdb <- pbdb %>% st_drop_geometry() %>% 
  mutate(mid_ma = mid_ma.x) %>% 
  dplyr::select(Name, bin_assignment, period, mid_ma) %>% 
  distinct(Name, period, bin_assignment, mid_ma)

# Mass and RBL data for extinct species and combine that with that of extant species from sections above
massrbl_porto <- read.delim("mass_&_RBL_porto.txt")
extinct <- left_join(pbdb,massrbl_porto,by=c("Name" = "species")) 
extant <- traits %>% dplyr::select(Name, RBL_original, Mass_original) %>% 
  mutate(period = "extant") %>% 
  mutate(mass = Mass_original, rbl = RBL_original) %>% 
  dplyr::select(Name, mass, rbl, period)
historical <- bind_rows(extant, extinct) 
historical_global <- historical %>% dplyr::select(Name, rbl, mass, period, bin_assignment, mid_ma) %>% distinct()

# Phylogenetic imputation of missing data using Rphylopars 
tree <- phylo[[1]]
historical_global <- historical_global %>% filter(Name %in% tree$tip.label)
imputation <- historical_global %>% mutate(species = Name) %>% dplyr::select(species, rbl, mass)
imputation <- imputation[imputation$species %in% tree$tip.label,]
imputation <- data.frame( species = imputation$species, rbl = imputation$rbl, mass = imputation$mass, stringsAsFactors = FALSE)
tree$edge.length[which(tree$edge.length==0)]=0.0001
fit <- Rphylopars::phylopars(trait_data = imputation, tree = tree)
imputed <- fit$trait_data

# Rejoin imputed dataset with original dataset
historical_global_imputed <- left_join(historical_global, imputed, by = c("Name" = "species")) %>% 
  mutate(rbl = case_when(!is.na(rbl.x)~rbl.x, TRUE ~rbl.y),
         mass = case_when(!is.na(mass.x)~mass.x, TRUE ~mass.y)) %>% dplyr::select(Name, rbl, mass, period, bin_assignment, mid_ma) %>% distinct() 
```

## Create trait space for each time period using funspace
```{r funspace periods}
# Create a dataframe with historical data for funspace
historical_funspace_df <- historical_global_imputed %>% filter(!is.na(rbl)) %>% 
  mutate(across(
    .cols = c(
      rbl, mass),
    .fns = ~log(.x), 
    .names = "{.col}_log"
  )) %>% 
  # rescale from 0 to 1
  mutate(across(
    .cols = ends_with("_log"),
    .fns = ~ scales::rescale(.x, to = c(0,1)),
    .names = "{.col}_scaled"
  )) 

historical_funspace_df$bin_assignment[is.na(historical_funspace_df$bin_assignment)] <- 200

# Add subfamily information (for reference only)
historical_funspace_df$genus <- sub("_.*","", historical_funspace_df$Name)
subfamily_lookup <- c(
  # Hesperocyoninae
  Hesperocyon="Hesperocyoninae", Otarocyon="Hesperocyoninae", Osbornodon="Hesperocyoninae", Archaeocyon="Hesperocyoninae", Mesocyon="Hesperocyoninae", 
  Cynodesmus="Hesperocyoninae", Enhydrocyon="Hesperocyoninae", Paraenhydrocyon="Hesperocyoninae", Cormocyon="Hesperocyoninae", Sunkahetanka="Hesperocyoninae",
  Ectopocynus = "Hesperocyoninae", Philotrox = "Hesperocyoninae",
  # Borophaginae
  Borophagus="Borophaginae", Epicyon="Borophaginae", Aelurodon="Borophaginae", Carpocyon="Borophaginae",Paratomarctus="Borophaginae", Protepicyon="Borophaginae",
  Cynarctus="Borophaginae", Cynarctoides="Borophaginae", Phlaocyon="Borophaginae", Metalopex="Borophaginae", Tomarctus = "Borophaginae", 
  Tephrocyon="Borophaginae", Euoplocyon="Borophaginae", Rhizocyon = "Borophaginae", Desmocyon = "Borophaginae", Metatomarctus = "Borophaginae", 
  Microtomarctus = "Borophaginae", Oxetocyon = "Borophaginae", Paracynarctus = "Borophaginae", Protomarctus = "Borophaginae", Psalidocyon = "Borophaginae",
  # Caninae
  Canis="Caninae", Vulpes="Caninae", Lycalopex="Caninae", Lycaon="Caninae", Cuon="Caninae", Urocyon="Caninae", Cerdocyon="Caninae", Chrysocyon="Caninae", 
  Speothos="Caninae", Atelocynus="Caninae", Nyctereutes="Caninae", Xenocyon="Caninae", Eucyon="Caninae", Leptocyon="Caninae", Protocyon="Caninae", Otocyon = "Caninae"
)
historical_funspace_df$subfamily <- subfamily_lookup[historical_funspace_df$genus]

# Add Time of Speciation (TS) and Time of Extinction (TE) data 
tste <- read.delim("TS_TE_Canidae.txt")
historical_funspace_df <-  left_join(historical_funspace_df, tste, by = c("Name" = "species"))

# Create new columns that specify which major canid evolutionary stage the species are present in
historical_funspace_df <- historical_funspace_df %>% 
  mutate(Emergence_of_Canidae = case_when(TE<=37.5 & TS >=33.7 ~"retained", TRUE~"not")) %>% 
  mutate(Decline_of_broadleaf_forest = case_when(TS>=28.5 & TE<=33.7 ~"retained", TRUE~"not")) %>% 
  mutate(Expansion_of_sparse_temperate_forest = case_when(TS>=25.7 & TE<=28.5~"retained", TRUE~"not")) %>% 
  mutate(Catgap = case_when(TS>=18.7 & TE<=25.7~"retained", TRUE~"not")) %>% 
  mutate(Catgap_filled = case_when(TS>=16.7& TE<=18.7 ~"retained", TRUE~"not")) %>% 
  mutate(Mid_miocene_climatic_optimum = case_when(TS>=13.8 & TE<=16.7~"retained", TRUE~"not")) %>% 
  mutate(Extinction_of_Hesperocyoninaes = case_when(TS>=9.8 & TE<=13.8~"retained", TRUE~"not")) %>% 
  mutate(C4_grassland_expansion = case_when(TS>=7.5 & TE<=9.8~"retained", TRUE~"not")) %>% 
  mutate(Maximum_diversity_of_horses = case_when(TS>=4.3 & TE<=7.5~"retained", TRUE~"not")) %>% 
  mutate(Isthmus_of_Panama = case_when(TS>=2.6& TE<=4.3 ~"retained", TRUE~"not")) %>% 
  mutate(Extinction_of_Borophaginaes = case_when(TS>=0.4 & TE<=2.6~"retained", TRUE~"not")) %>% 
  mutate(Megafauna_extinction = case_when(TS>=0 & TE<=0.4~"retained", TRUE~"not")) %>% 
  mutate(Current = case_when(period == "extant" ~"extant", TRUE~"extinct")) %>% 
  mutate(RBL = rbl_log_scaled, Mass = mass_log_scaled)

# Construct trait-space for all time periods using funspace
historical_funspace_df_num <- historical_funspace_df %>% dplyr::select(RBL, Mass)
historical_trait_space_funspace <- funspace(x = historical_funspace_df_num, n_divisions = 300)
summary(historical_trait_space_funspace)
plot(x = historical_trait_space_funspace, type = "global", quant.plot = TRUE, axis.title.x = "RBL", axis.title.y = "Mass")
historical_fric <- historical_trait_space_funspace$FD$global$FRich
historical_fdiv <- historical_trait_space_funspace$FD$global$FDiv

# Compare extant with historical all-time trait-space
historical_trait_space_funspace_extant <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Current`, n_divisions = 300)
plot(x = historical_trait_space_funspace_extant, type = "groups", which.group = "extant",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_extant)
Extant_fric <- historical_trait_space_funspace_extant$FD$groups$FRich["extant"]
Extant_fdiv <- historical_trait_space_funspace_extant$FD$groups$FDiv["extant"]
Extinct_fric <- historical_trait_space_funspace_extant$FD$groups$FRich["extinct"]
Extinct_fdiv <- historical_trait_space_funspace_extant$FD$groups$FDiv["extinct"]
  
extant_results <- tibble::tibble(
  Scenario = c("All-time", "Extant", "Extinct"),
  FRic = c(historical_fric, Extant_fric, Extinct_fric),
  FDiv = c(historical_fdiv, Extant_fdiv, Extinct_fdiv))

# Construct trait-space for all time periods individually using funspace
  # Emergence_of_Canidae
historical_trait_space_funspace_Emergence_of_Canidae <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Emergence_of_Canidae`, n_divisions = 300)
plot(x = historical_trait_space_funspace_Emergence_of_Canidae, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_Emergence_of_Canidae)
Emergence_of_Canidae_fric <- historical_trait_space_funspace_Emergence_of_Canidae$FD$groups$FRich["retained"]
Emergence_of_Canidae_fdiv <- historical_trait_space_funspace_Emergence_of_Canidae$FD$groups$FDiv["retained"]
  
  # Decline_of_broadleaf_forest
historical_trait_space_funspace_Decline_of_broadleaf_forest <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Decline_of_broadleaf_forest`, n_divisions = 300)
plot(x = historical_trait_space_funspace_Decline_of_broadleaf_forest, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_Decline_of_broadleaf_forest)
Decline_of_broadleaf_forest_fric <- historical_trait_space_funspace_Decline_of_broadleaf_forest$FD$groups$FRich["retained"]
Decline_of_broadleaf_forest_fdiv <- historical_trait_space_funspace_Decline_of_broadleaf_forest$FD$groups$FDiv["retained"]

  # Expansion_of_sparse_temperate_forest
historical_trait_space_funspace_Expansion_of_sparse_temperate_forest <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Expansion_of_sparse_temperate_forest`, n_divisions = 300)
plot(x = historical_trait_space_funspace_Expansion_of_sparse_temperate_forest, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_Expansion_of_sparse_temperate_forest)
Expansion_of_sparse_temperate_forest_fric <- historical_trait_space_funspace_Expansion_of_sparse_temperate_forest$FD$groups$FRich["retained"]
Expansion_of_sparse_temperate_forest_fdiv <- historical_trait_space_funspace_Expansion_of_sparse_temperate_forest$FD$groups$FDiv["retained"]
  
  # Catgap
historical_trait_space_funspace_Catgap <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Catgap`, n_divisions = 300)
plot(x = historical_trait_space_funspace_Catgap, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_Catgap)
Catgap_fric <- historical_trait_space_funspace_Catgap$FD$groups$FRich["retained"]
Catgap_fdiv <- historical_trait_space_funspace_Catgap$FD$groups$FDiv["retained"]

  # Catgap_filled
historical_trait_space_funspace_Catgap_filled <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Catgap_filled`, n_divisions = 300)
plot(x = historical_trait_space_funspace_Catgap_filled, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_Catgap_filled)
Catgap_filled_fric <- historical_trait_space_funspace_Catgap_filled$FD$groups$FRich["retained"]
Catgap_filled_fdiv <- historical_trait_space_funspace_Catgap_filled$FD$groups$FDiv["retained"]
  
  # Mid_miocene_climatic_optimum
historical_trait_space_funspace_Mid_miocene_climatic_optimum <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Mid_miocene_climatic_optimum`, n_divisions = 300)
plot(x = historical_trait_space_funspace_Mid_miocene_climatic_optimum, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_Mid_miocene_climatic_optimum)
Mid_miocene_climatic_optimum_fric <- historical_trait_space_funspace_Mid_miocene_climatic_optimum$FD$groups$FRich["retained"]
Mid_miocene_climatic_optimum_fdiv <- historical_trait_space_funspace_Mid_miocene_climatic_optimum$FD$groups$FDiv["retained"]

  # Extinction_of_Hesperocyoninaes
historical_trait_space_funspace_Extinction_of_Hesperocyoninaes <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Extinction_of_Hesperocyoninaes`, n_divisions = 300)
plot(x = historical_trait_space_funspace_Extinction_of_Hesperocyoninaes, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_Extinction_of_Hesperocyoninaes)
Extinction_of_Hesperocyoninaes_fric <- historical_trait_space_funspace_Extinction_of_Hesperocyoninaes$FD$groups$FRich["retained"]
Extinction_of_Hesperocyoninaes_fdiv <- historical_trait_space_funspace_Extinction_of_Hesperocyoninaes$FD$groups$FDiv["retained"]
  
  # C4_grassland_expansion
historical_trait_space_funspace_C4_grassland_expansion <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`C4_grassland_expansion`, n_divisions = 300)
plot(x = historical_trait_space_funspace_C4_grassland_expansion, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_C4_grassland_expansion)
C4_grassland_expansion_fric <- historical_trait_space_funspace_C4_grassland_expansion$FD$groups$FRich["retained"]
C4_grassland_expansion_fdiv <- historical_trait_space_funspace_C4_grassland_expansion$FD$groups$FDiv["retained"]

  # Maximum_diversity_of_horses
historical_trait_space_funspace_Maximum_diversity_of_horses <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Maximum_diversity_of_horses`, n_divisions = 300)
plot(x = historical_trait_space_funspace_Maximum_diversity_of_horses, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_Maximum_diversity_of_horses)
Maximum_diversity_of_horses_fric <- historical_trait_space_funspace_Maximum_diversity_of_horses$FD$groups$FRich["retained"]
Maximum_diversity_of_horses_fdiv <- historical_trait_space_funspace_Maximum_diversity_of_horses$FD$groups$FDiv["retained"]
  
  # Isthmus_of_Panama
historical_trait_space_funspace_Isthmus_of_Panama <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Isthmus_of_Panama`, n_divisions = 300)
plot(x = historical_trait_space_funspace_Isthmus_of_Panama, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_Isthmus_of_Panama)
Isthmus_of_Panama_fric <- historical_trait_space_funspace_Isthmus_of_Panama$FD$groups$FRich["retained"]
Isthmus_of_Panama_fdiv <- historical_trait_space_funspace_Isthmus_of_Panama$FD$groups$FDiv["retained"]

  # Extinction_of_Borophaginaes
historical_trait_space_funspace_Extinction_of_Borophaginaes <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Extinction_of_Borophaginaes`, n_divisions = 300)
plot(x = historical_trait_space_funspace_Extinction_of_Borophaginaes, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_Extinction_of_Borophaginaes)
Extinction_of_Borophaginaes_fric <- historical_trait_space_funspace_Extinction_of_Borophaginaes$FD$groups$FRich["retained"]
Extinction_of_Borophaginaes_fdiv <- historical_trait_space_funspace_Extinction_of_Borophaginaes$FD$groups$FDiv["retained"]
  
  # Megafauna_extinction
historical_trait_space_funspace_Megafauna_extinction <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Megafauna_extinction`, n_divisions = 300)
plot(x = historical_trait_space_funspace_Megafauna_extinction, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_Megafauna_extinction)
Megafauna_extinction_fric <- historical_trait_space_funspace_Megafauna_extinction$FD$groups$FRich["retained"]
Megafauna_extinction_fdiv <- historical_trait_space_funspace_Megafauna_extinction$FD$groups$FDiv["retained"]

# Collate all results
historical_results <- tibble::tibble(
  Scenario = c("Global", "Emergence of Canidae", "Decline of broadleaf forest", "Expansion of sparse temperate forest", "Catgap", "Catgap", "Mid miocene climatic optimum",
               "Extinction of Hesperocyoninaes", "C4 grassland expansion", "Maximum diversity of horses", "Isthmus of Panama", "Extinction of Borophaginaes", "Megafauna extinction"),
  FRic = c(historical_fric, Emergence_of_Canidae_fric, Decline_of_broadleaf_forest_fric, Expansion_of_sparse_temperate_forest_fric, Catgap_fric, Catgap_filled_fric,
           Mid_miocene_climatic_optimum_fric, Extinction_of_Hesperocyoninaes_fric, C4_grassland_expansion_fric, Maximum_diversity_of_horses_fric, Isthmus_of_Panama_fric,
           Extinction_of_Borophaginaes_fric, Megafauna_extinction_fric),
  FDiv = c(historical_fdiv, Emergence_of_Canidae_fdiv, Decline_of_broadleaf_forest_fdiv, Expansion_of_sparse_temperate_forest_fdiv, Catgap_fdiv, Catgap_filled_fdiv,
           Mid_miocene_climatic_optimum_fdiv, Extinction_of_Hesperocyoninaes_fdiv, C4_grassland_expansion_fdiv, Maximum_diversity_of_horses_fdiv, Isthmus_of_Panama_fdiv,
           Extinction_of_Borophaginaes_fdiv, Megafauna_extinction_fdiv))

historical_results
```

## Plot results
```{r plot historical}
# All periods
events_list <- list(
  Emergence_of_Canidae = historical_trait_space_funspace_Emergence_of_Canidae,
  Decline_of_broadleaf_forest = historical_trait_space_funspace_Decline_of_broadleaf_forest,
  Expansion_of_sparse_temperate_forest = historical_trait_space_funspace_Expansion_of_sparse_temperate_forest,
  Catgap = historical_trait_space_funspace_Catgap,
  Catgap_filled = historical_trait_space_funspace_Catgap_filled,
  Mid_miocene_climatic_optimum = historical_trait_space_funspace_Mid_miocene_climatic_optimum,
  Extinction_of_Hesperocyoninaes = historical_trait_space_funspace_Extinction_of_Hesperocyoninaes,
  C4_grassland_expansion = historical_trait_space_funspace_C4_grassland_expansion,
  Maximum_diversity_of_horses = historical_trait_space_funspace_Maximum_diversity_of_horses,
  Isthmus_of_Panama = historical_trait_space_funspace_Isthmus_of_Panama,
  Extinction_of_Borophaginaes = historical_trait_space_funspace_Extinction_of_Borophaginaes,
  Megafauna_extinction = historical_trait_space_funspace_Megafauna_extinction
)

par(mfrow=c(3,4))

for (nm in names(events_list)) {

  plot(
    x = events_list[[nm]],
    type = "groups",
    which.group = "retained",
    quant.plot = TRUE,
    globalContour = TRUE,
    pnt = TRUE,
    pnt.cex = 0.1,
    pnt.col = rgb(0.2,0.8,0.1,alpha=0.2),
    axis.title.x = "RBL",
    axis.title.y = "Mass",
    axis.title.cex = 0.8,
    axis.cex = 0.5
  )

  title(gsub("_"," ",nm), cex.main = 1.1)
}
```