---
title: "canid_final"
author: "Dorothy Woo"
date: "2026-07-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Functional Diversity of Canids (Canidae) Under Projected Anthropogenic Extinction as Determined by Functional Trait-space Analysis

## RSJZ0, UCL MRes Masters Dissertation
## Supervisors: Dr Lucas Porto, Dr Natalie Cooper (Natural History Museum, London)

The code below was used to answer study objects in the dissertation

# Project set up
## Load packages
```{r load packages, message=FALSE, warning=FALSE}
library(devtools)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(patchwork)
library(ggthemes)
library(lme4)
library("lmerTest")
library(emmeans)
library(vegan)
library(sf)
library(stringr)
library(terra)
library(purrr)
library(missMDA) 
library(FactoMineR) 
library(nnet) 
library(randomForest)
library(viridis)
library(factoextra)
library(knitr)
library(patchwork)
library(tidyr)
library(magick)
library(grid)
library(ggforce)
library(png)
library(Claddis)  
library(geoscale) 
library(dismo)  
library(ape)
library(tree)
library(funspace)
library(traitdata)
library(reticulate)
library(iucnsim)
library(rredlist)
library(corrplot)
library(psych)
library(scales)
library(plotly)
library(ggrepel)
library(geometry)
library(gridExtra)
library(svglite)
library(geiger)
library(cowplot)
library(tibble)
library(sp)
library(zoo)
library(gganimate)
library(phytools)
library(palaeoverse)
library(Rphylopars)
library(hypervolume)
library(rgl)
library(FNN)
library(picante)
library(lubridate)
```

# Data preparation
## Load phylogeny 
```{r load phylogeny,message=FALSE,warning=FALSE}
# Get all species name
canidae_species_names <- read.csv(
  "canidae_SFaurby.csv",
  stringsAsFactors = FALSE
) 
canidae_species_names <- paste(canidae_species_names$Genus, canidae_species_names$Species, sep = "_")

# Load phylogenetic tree
phylo <- ape::read.nexus("SFaurby_phylogeny.nex")   

# Trim tree to contain only Canidae
canidae_phylo <- sapply(phylo, function(tr) {
  any(tr$tip.label %in% canidae_species_names)
})
phylo_canidae <- phylo[canidae_phylo]
phylo_canidae <- lapply(phylo_canidae, function(tr) {
  keep <- intersect(tr$tip.label, canidae_species_names)
  drop.tip(tr, setdiff(tr$tip.label, keep))
})
class(phylo_canidae) <- "multiPhylo"
```

## Load data with IUCN status
```{r iucn, message=FALSE}
iucn_status <- read_csv("iucn_status.csv")
```

## Load PanTHERIA data 
```{r load pantheria data, warning=FALSE, message=FALSE}
# pantheria
data("pantheria")
pantheria <- pantheria %>% filter(Family=="Canidae") %>% mutate(Name=paste0(Genus,"_",Species))
pantheria <- pantheria[, colMeans(!is.na(pantheria)) > 0.5] # retain columns with <50% missing data
```

## Load RBL dataset
```{r rbl data, warning=FALSE, message=FALSE}
rbl <- read_csv("rbl_martina.csv") %>% mutate(Name = TaxonName) %>% dplyr::select(Name, RBL)
```

## Compile all traits
```{r all traits data,message=FALSE}
# Select traits needed
all_data <- list(
  iucn_status,
  pantheria,
  rbl
)

traits <- purrr::reduce(all_data, full_join, by = "Name") 

write_csv(traits, "alltraits.csv")
```

## Reload data after manually changing names that are duplicated
```{r trait data processing, message=FALSE, warning=FALSE}
# Reload trait data after manual fix
traits <- read_csv("selected_traits.csv")
## In this excel, I've also defined all extinction scenarios manually 
```

## Log transform all traits 
```{r transform data}
traits <- traits %>% 
  # log-transform
  mutate(across(
    .cols = c(
      RBL_original,
      Mass_original),
    .fns = ~log(.x), 
    .names = "{.col}_log"
  )) %>% 
  # rescale from 0 to 1
  mutate(across(
    .cols = ends_with("_log"),
    .fns = ~ scales::rescale(.x, to = c(0,1)),
    .names = "{.col}_scaled"
  )) 
```
## Define IUCN colors 
```{r iucn colors}
# Define IUCN colors
iucn_cols <- c(
  "LC" = "grey80",
  "NT" = "darkturquoise",
  "EN" = "orange",
  "CR" = "red")
```

## Define continent colours
```{r cont colors}
# Define continent colors
continent_cols <- c(
  "Global" = "black",
  "South America" = "hotpink",
  "North America" = "lightblue",
  "Africa" = "lightgreen",
  "Eurasia" = "yellow",
  "Asia" = "orange",
  "Europe" = "purple")
```


## Prepare funspace df
```{r extinction scenarios}
funspace_df <- traits %>% dplyr::select(
  Name, IUCN_status, continents, RBL_original_log_scaled, Mass_original_log_scaled,
  Extinction_CR_EN_NT, Extinction_SA_NT, Extinction_NA_NT, Extinction_Africa_NT, Extinction_Eurasia_NT) %>% 
  mutate(RBL = RBL_original_log_scaled, Mass = Mass_original_log_scaled) 
```

# Research Objective 1: How do extinction events affect FD globally
## Section 1: Global extinction of threatened and near-threatened species
```{r funspace global}
funspace_df_num <- funspace_df %>% dplyr::select(RBL, Mass)

# Create trait space
traitspace <- funspace(x = funspace_df_num, n_divisions = 300)
summary(traitspace)
plot(x = traitspace, type = "global", quant.plot = TRUE, arrows = TRUE, arrows.length = 0.6, axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")

# Global extinction of CR EN NT
traitspace_CR_EN_NT <-  funspace(x = funspace_df_num, group.vec = funspace_df$Extinction_CR_EN_NT, n_divisions = 300)
summary(traitspace_CR_EN_NT)
plot(x = traitspace_CR_EN_NT, type = "groups", quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")

# Compile results
global_fric_original <- traitspace$FD$global$FRich
global_fdiv_original <- traitspace$FD$global$FDiv
global_extinction_fric <- traitspace_CR_EN_NT$FD$groups$FRich["retained"]
global_extinction_fdiv <- traitspace_CR_EN_NT$FD$groups$FDiv["retained"]
global_fric_change <- 100*(global_extinction_fric-global_fric_original)/global_fric_original
globa_fdiv_change <- 100*(global_extinction_fdiv-global_fdiv_original)/global_fdiv_original

global_results <- tibble::tibble(
  Scenario = c("Global", "Removing CR, EN, NT"),
  FRic = c(global_fric_original, global_extinction_fric),
  FRic_change_percent = c("",global_fric_change),
  FDiv = c(global_fdiv_original, global_extinction_fdiv),
  FDiv_change_percent = c("", globa_fdiv_change)
)

global_results
```
## Section 2: Continental extinction of threatened and near-threatened species
```{r funspace continent}
# Create trait space for each continent
traitspace_continent <-  funspace(x = funspace_df_num, PCs=c(1,2), group.vec = funspace_df$continents, n_divisions = 300)
plot(x = traitspace_continent, type = "groups", quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(traitspace_continent)

par(mfrow=c(2,2))
# Removing SA NT
traitspace_continent_SA <-  funspace(x = funspace_df_num, group.vec = funspace_df$Extinction_SA_NT, n_divisions = 300)
plot(x = traitspace_continent_SA, type = "groups", which.group = "retained", quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass", main = "South America")
title("South America only", cex.main = 1)
summary(traitspace_continent_SA)

# Removing North America NT
traitspace_continent_NA <-  funspace(x = funspace_df_num, group.vec = funspace_df$Extinction_NA_NT, n_divisions = 300)
plot(x = traitspace_continent_NA, type = "groups", which.group = "retained", quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
title("North America only", cex.main = 1)
summary(traitspace_continent_NA)

# Removing Africa NT
traitspace_continent_Africa <-  funspace(x = funspace_df_num, group.vec = funspace_df$Extinction_Africa_NT, n_divisions = 300)
plot(x = traitspace_continent_Africa, type = "groups", which.group = "retained", quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
title("Africa only", cex.main = 1)
summary(traitspace_continent_Africa)

# Removing Eurasia NT
traitspace_continent_Eurasia <-  funspace(x = funspace_df_num, group.vec = funspace_df$Extinction_Eurasia_NT, n_divisions = 300)
plot(x = traitspace_continent_Eurasia, type = "groups", which.group = "retained", quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
title("Eurasia only", cex.main = 1)
summary(traitspace_continent_Eurasia)

# Compile results
sa_fric <- traitspace_continent_SA$FD$groups$FRich["retained"]
sa_fdiv <- traitspace_continent_SA$FD$groups$FDiv["retained"]

na_fric <- traitspace_continent_NA$FD$groups$FRich["retained"]
na_fdiv <- traitspace_continent_NA$FD$groups$FDiv["retained"]

africa_fric <- traitspace_continent_Africa$FD$groups$FRich["retained"]
africa_fdiv <- traitspace_continent_Africa$FD$groups$FDiv["retained"]

eurasia_fric <- traitspace_continent_Eurasia$FD$groups$FRich["retained"]
eurasia_fdiv <- traitspace_continent_Eurasia$FD$groups$FDiv["retained"]

sa_fric_change <- 100*(sa_fric-global_fric_original)/global_fric_original
na_fric_change <- 100*(na_fric-global_fric_original)/global_fric_original
africa_fric_change <- 100*(africa_fric-global_fric_original)/global_fric_original
eurasia_fric_change <- 100*(eurasia_fric-global_fric_original)/global_fric_original

sa_fdiv_change <- 100*(sa_fdiv-global_fdiv_original)/global_fdiv_original
na_fdiv_change <- 100*(na_fdiv-global_fdiv_original)/global_fdiv_original
africa_fdiv_change <- 100*(africa_fdiv-global_fdiv_original)/global_fdiv_original
eurasia_fdiv_change <- 100*(eurasia_fdiv-global_fdiv_original)/global_fdiv_original

continental_results <- tibble::tibble(
  Continent = c("Global", "South America", "North America",  "Africa", "Eurasia"),
  FRic = c(global_fric_original, sa_fric, na_fric, africa_fric, eurasia_fric),
  FRic_change_percent = c("",sa_fric_change, na_fric_change, 
                          africa_fric_change,eurasia_fric_change),
  FDiv = c(global_fdiv_original, sa_fdiv, na_fdiv, africa_fdiv, eurasia_fdiv),
  FDiv_change_percent = c("",sa_fdiv_change, na_fdiv_change, 
                          africa_fdiv_change,eurasia_fdiv_change),
)

continental_results
```

# Research Objective 2: How do extinction events affect FD continentally 
```{r continent separate results}
# Comparing FRich for each continent after removing CR/EN/NT species 
par(mfrow = c(2,2))
  # South America
df_temp <- funspace_df %>% filter(continents == "South America")
df_temp_num <- df_temp %>% dplyr::select(RBL, Mass)
temp_funspace <- funspace(x = df_temp_num, group.vec = df_temp$Extinction_SA_NT, n_divisions = 300)
summary(temp_funspace)
SA_cfrich_original <- temp_funspace$FD$global$FRich
SA_cfrich_new <- temp_funspace$FD$groups$FRich
SA_cfdiv_original <- temp_funspace$FD$global$FDiv
SA_cfdiv_new <- temp_funspace$FD$groups$FDiv
plot(x = temp_funspace, type = "groups", which.group = "retained", quant.plot = TRUE, globalContour = T, colors = c("cadetblue1", "blue"), pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass") 
title("South America", cex.main = 1)

  # North America
df_temp <- funspace_df %>% filter(continents == "North America")
df_temp_num <- df_temp %>% dplyr::select(RBL, Mass)
temp_funspace <- funspace(x = df_temp_num, group.vec = df_temp$Extinction_NA_NT, n_divisions = 300)
summary(temp_funspace)
NA_cfrich_original <- temp_funspace$FD$global$FRich
NA_cfrich_new <- temp_funspace$FD$groups$FRich
NA_cfdiv_original <- temp_funspace$FD$global$FDiv
NA_cfdiv_new <- temp_funspace$FD$groups$FDiv
plot(x = temp_funspace, type = "groups", which.group = "retained", quant.plot = TRUE, globalContour = T, colors = c("cadetblue1", "blue"), pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
title("North America", cex.main = 1)


  # Africa 
df_temp <- funspace_df %>% filter(continents == "Africa")
df_temp_num <- df_temp %>% dplyr::select(RBL, Mass)
temp_funspace <- funspace(x = df_temp_num, group.vec = df_temp$Extinction_Africa_NT, n_divisions = 300)
summary(temp_funspace)
Africa_cfrich_original <- temp_funspace$FD$global$FRich
Africa_cfrich_new <- temp_funspace$FD$groups$FRich
Africa_cfdiv_original <- temp_funspace$FD$global$FDiv
Africa_cfdiv_new <- temp_funspace$FD$groups$FDiv
plot(x = temp_funspace, type = "groups", which.group = "retained", quant.plot = TRUE, globalContour = T, colors = c("cadetblue1", "blue"), pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
title("Africa", cex.main = 1)

  # Eurasia
df_temp <- funspace_df %>% filter(continents == "Eurasia")
df_temp_num <- df_temp %>% dplyr::select(RBL, Mass)
temp_funspace <- funspace(x = df_temp_num, group.vec = df_temp$Extinction_Eurasia_NT, n_divisions = 300)
summary(temp_funspace)
Eurasia_cfrich_original <- temp_funspace$FD$global$FRich
Eurasia_cfrich_new <- temp_funspace$FD$groups$FRich
Eurasia_cfdiv_original <- temp_funspace$FD$global$FDiv
Eurasia_cfdiv_new <- temp_funspace$FD$groups$FDiv
plot(x = temp_funspace, type = "groups", which.group = "retained", quant.plot = TRUE, globalContour = T, colors = c("cadetblue1", "blue"), pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
title("Eurasia", cex.main = 1)

# FRich percentage change results
SA_cfrich_new <- as.numeric(SA_cfrich_new["retained"])
SA_cfdiv_new <- as.numeric(SA_cfdiv_new["retained"])
SA_cfrich_percent <- ((SA_cfrich_new-SA_cfrich_original)/SA_cfrich_original)*100
SA_cfdiv_percent <- ((SA_cfdiv_new-SA_cfdiv_original)/SA_cfdiv_original)*100

NA_cfrich_new <- as.numeric(NA_cfrich_new["retained"])
NA_cfdiv_new <- as.numeric(NA_cfdiv_new["retained"])
NA_cfrich_percent <- ((NA_cfrich_new-NA_cfrich_original)/NA_cfrich_original)*100
NA_cfdiv_percent <- ((NA_cfdiv_new-NA_cfdiv_original)/NA_cfdiv_original)*100

Africa_cfrich_new <- as.numeric(Africa_cfrich_new["retained"])
Africa_cfdiv_new <- as.numeric(Africa_cfdiv_new["retained"])
Africa_cfrich_percent <- ((Africa_cfrich_new-Africa_cfrich_original)/Africa_cfrich_original)*100
Africa_cfdiv_percent <- ((Africa_cfdiv_new-Africa_cfdiv_original)/Africa_cfdiv_original)*100

Eurasia_cfrich_new <- as.numeric(Eurasia_cfrich_new["retained"])
Eurasia_cfdiv_new <- as.numeric(Eurasia_cfdiv_new["retained"])
Eurasia_cfrich_percent <- ((Eurasia_cfrich_new-Eurasia_cfrich_original)/Eurasia_cfrich_original)*100
Eurasia_cfdiv_percent <- ((Eurasia_cfdiv_new-Eurasia_cfdiv_original)/Eurasia_cfdiv_original)*100

# Compile results
continent_standalone_results <- tibble::tibble(
  Continent = c("South America", "South America", "North America", "North America", "Africa","Africa", "Eurasia", "Eurasia"),
  Scenario = c("Original", "Post-extinction", "Original", "Post-extinction", "Original", "Post-extinction", "Original", "Post-extinction"),
  FRic = c(SA_cfrich_original, SA_cfrich_new, NA_cfrich_original, NA_cfrich_new, 
           Africa_cfrich_original, Africa_cfrich_new, Eurasia_cfrich_original, Eurasia_cfrich_new),
  Change_in_FRic = c("-", SA_cfrich_percent, "-", NA_cfrich_percent, "-", Africa_cfrich_percent, "-", Eurasia_cfrich_percent),
  FDiv = c(SA_cfdiv_original, SA_cfdiv_new, NA_cfdiv_original, NA_cfdiv_new, 
           Africa_cfdiv_original, Africa_cfdiv_new, Eurasia_cfdiv_original, Eurasia_cfdiv_new),
  Change_in_FDiv = c("-", SA_cfdiv_percent, "-", NA_cfdiv_percent, "-", Africa_cfdiv_percent, "-", Eurasia_cfdiv_percent))

continent_standalone_results
```

# Research Objective 3: How do contemporary, anthropogenic extinction events compare to past historical events
## Use PBDB database to get continent and time information
```{r pbdb, warning=FALSE, message=FALSE}
pbdb <- read_csv("pbdb_data.csv")
pbdb <- pbdb %>% dplyr::select(Name, lng, lat, max_ma, min_ma) %>% mutate(mid_ma = (max_ma+min_ma)/2)

# convert longitude and latitude points to sf points 
pbdb_sfpts <- st_as_sf(pbdb, coords = c("lng","lat"), crs = 4326)
# overlay with continent (check iucn geographic chunk)
pbdb_sfpts <- st_transform(pbdb_sfpts, st_crs(continents))
# add new continent column
pbdb <- st_join(pbdb_sfpts, continents["continent"])

# assign a period for each species
bins <- time_bins(scale = "international ages")
pbdb <- bin_time(
  occdf = pbdb,
  bins = bins,
  min_ma = "min_ma",
  max_ma = "max_ma",
  method = "majority"
)
bins_name <- dplyr::select(bins, bin, interval_name) 
pbdb <- left_join(pbdb, bins, by=c("bin_assignment" = "bin"))
pbdb <- pbdb %>% mutate(period = interval_name) 
pbdb <- pbdb %>% st_drop_geometry() %>% 
  mutate(mid_ma = mid_ma.x) %>% 
  dplyr::select(Name, continent, bin_assignment, period, mid_ma) %>% 
  distinct(Name, continent, period, bin_assignment, mid_ma)

# Mass and RBL data
massrbl_porto <- read.delim("mass_&_RBL_porto.txt")
supp_rbl <- read_csv("supp_rbl.csv") %>% dplyr::select(species, `blade/m1`)
extinct <- left_join(pbdb,massrbl_porto,by=c("Name" = "species")) 
extinct <- left_join(extinct,supp_rbl,by=c("Name" = "species")) 
extant <- traits %>% dplyr::select(Name, RBL_original, Mass_original, continents) %>% 
  mutate(period = "extant") %>% 
  separate_rows(continents, sep = ", ") %>% 
  mutate(mass = Mass_original, rbl = RBL_original, continent = continents) %>% 
  dplyr::select(Name, mass, rbl, continent, period)
historical <- bind_rows(extant, extinct) 
historical_global <- historical %>% dplyr::select(Name, rbl, mass, period, bin_assignment, mid_ma) %>% distinct()

# since some mass and rbl data is missing for extinct species, use rphylopars to imputate 
tree <- phylo[[1]]
historical_global <- historical_global %>% filter(Name %in% tree$tip.label)
imputation <- historical_global %>% mutate(species = Name) %>% dplyr::select(species, rbl, mass)
imputation <- imputation[imputation$species %in% tree$tip.label,]
imputation <- data.frame( species = imputation$species, rbl = imputation$rbl, mass = imputation$mass, stringsAsFactors = FALSE)
tree$edge.length[which(tree$edge.length==0)]=0.0001
fit <- Rphylopars::phylopars(trait_data = imputation, tree = tree)
imputed <- fit$trait_data

# rejoin imputed dataset with original dataset
historical_global_imputed <- left_join(historical_global, imputed, by = c("Name" = "species")) %>% 
  mutate(rbl = case_when(!is.na(rbl.x)~rbl.x, TRUE ~rbl.y),
         mass = case_when(!is.na(mass.x)~mass.x, TRUE ~mass.y)) %>% dplyr::select(Name, rbl, mass, period, bin_assignment, mid_ma) %>% distinct() 
```

## Create trait space for each time period funspace
```{r funspace periods}
historical_funspace_df <- historical_global_imputed %>% filter(!is.na(rbl)) %>% 
  mutate(across(
    .cols = c(
      rbl, mass),
    .fns = ~log(.x), 
    .names = "{.col}_log"
  )) %>% 
  # rescale from 0 to 1
  mutate(across(
    .cols = ends_with("_log"),
    .fns = ~ scales::rescale(.x, to = c(0,1)),
    .names = "{.col}_scaled"
  )) 

historical_funspace_df$bin_assignment[is.na(historical_funspace_df$bin_assignment)] <- 200

# add subfamily information
historical_funspace_df$genus <- sub("_.*","", historical_funspace_df$Name)
subfamily_lookup <- c(
  # Hesperocyoninae
  Hesperocyon="Hesperocyoninae", Otarocyon="Hesperocyoninae", Osbornodon="Hesperocyoninae", Archaeocyon="Hesperocyoninae", Mesocyon="Hesperocyoninae", 
  Cynodesmus="Hesperocyoninae", Enhydrocyon="Hesperocyoninae", Paraenhydrocyon="Hesperocyoninae", Cormocyon="Hesperocyoninae", Sunkahetanka="Hesperocyoninae",
  Ectopocynus = "Hesperocyoninae", Philotrox = "Hesperocyoninae",
  # Borophaginae
  Borophagus="Borophaginae", Epicyon="Borophaginae", Aelurodon="Borophaginae", Carpocyon="Borophaginae",Paratomarctus="Borophaginae", Protepicyon="Borophaginae",
  Cynarctus="Borophaginae", Cynarctoides="Borophaginae", Phlaocyon="Borophaginae", Metalopex="Borophaginae", Tomarctus = "Borophaginae", 
  Tephrocyon="Borophaginae", Euoplocyon="Borophaginae", Rhizocyon = "Borophaginae", Desmocyon = "Borophaginae", Metatomarctus = "Borophaginae", 
  Microtomarctus = "Borophaginae", Oxetocyon = "Borophaginae", Paracynarctus = "Borophaginae", Protomarctus = "Borophaginae", Psalidocyon = "Borophaginae",
  # Caninae
  Canis="Caninae", Vulpes="Caninae", Lycalopex="Caninae", Lycaon="Caninae", Cuon="Caninae", Urocyon="Caninae", Cerdocyon="Caninae", Chrysocyon="Caninae", 
  Speothos="Caninae", Atelocynus="Caninae", Nyctereutes="Caninae", Xenocyon="Caninae", Eucyon="Caninae", Leptocyon="Caninae", Protocyon="Caninae", Otocyon = "Caninae"
)
historical_funspace_df$subfamily <- subfamily_lookup[historical_funspace_df$genus]

# also add ts te data
tste <- read.delim("TS_TE_Canidae.txt")
historical_funspace_df <-  left_join(historical_funspace_df, tste, by = c("Name" = "species"))

# periods to look at
historical_funspace_df <- historical_funspace_df %>% 
  mutate(Emergence_of_Canidae = case_when(TE<=37.5 & TS >=33.7 ~"retained", TRUE~"not")) %>% 
  mutate(Decline_of_broadleaf_forest = case_when(TS>=28.5 & TE<=33.7 ~"retained", TRUE~"not")) %>% 
  mutate(Expansion_of_sparse_temperate_forest = case_when(TS>=25.7 & TE<=28.5~"retained", TRUE~"not")) %>% 
  mutate(Catgap = case_when(TS>=18.7 & TE<=25.7~"retained", TRUE~"not")) %>% 
  mutate(Catgap_filled = case_when(TS>=16.7& TE<=18.7 ~"retained", TRUE~"not")) %>% 
  mutate(Mid_miocene_climatic_optimum = case_when(TS>=13.8 & TE<=16.7~"retained", TRUE~"not")) %>% 
  mutate(Extinction_of_Hesperocyoninaes = case_when(TS>=9.8 & TE<=13.8~"retained", TRUE~"not")) %>% 
  mutate(C4_grassland_expansion = case_when(TS>=7.5 & TE<=9.8~"retained", TRUE~"not")) %>% 
  mutate(Maximum_diversity_of_horses = case_when(TS>=4.3 & TE<=7.5~"retained", TRUE~"not")) %>% 
  mutate(Isthmus_of_Panama = case_when(TS>=2.6& TE<=4.3 ~"retained", TRUE~"not")) %>% 
  mutate(Extinction_of_Borophaginaes = case_when(TS>=0.4 & TE<=2.6~"retained", TRUE~"not")) %>% 
  mutate(Megafauna_extinction = case_when(TS>=0 & TE<=0.4~"retained", TRUE~"not")) %>% 
  mutate(Current = case_when(period == "extant" ~"extant", TRUE~"extinct")) %>% 
  mutate(RBL = rbl_log_scaled, Mass = mass_log_scaled)

# run funspace
historical_funspace_df_num <- historical_funspace_df %>% dplyr::select(RBL, Mass)
historical_trait_space_funspace <- funspace(x = historical_funspace_df_num, n_divisions = 300)
summary(historical_trait_space_funspace)
plot(x = historical_trait_space_funspace, type = "global", quant.plot = TRUE, axis.title.x = "RBL", axis.title.y = "Mass")
historical_fric <- historical_trait_space_funspace$FD$global$FRich
historical_fdiv <- historical_trait_space_funspace$FD$global$FDiv

# past vs present (all-time)
historical_trait_space_funspace_extant <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Current`, n_divisions = 300)
plot(x = historical_trait_space_funspace_extant, type = "groups", which.group = "extant",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_extant)
Extant_fric <- historical_trait_space_funspace_extant$FD$groups$FRich["extant"]
Extant_fdiv <- historical_trait_space_funspace_extant$FD$groups$FDiv["extant"]
Extinct_fric <- historical_trait_space_funspace_extant$FD$groups$FRich["extinct"]
Extinct_fdiv <- historical_trait_space_funspace_extant$FD$groups$FDiv["extinct"]
  
extant_results <- tibble::tibble(
  Scenario = c("All-time", "Extant", "Extinct"),
  FRic = c(historical_fric, Extant_fric, Extinct_fric),
  FDiv = c(historical_fdiv, Extant_fdiv, Extinct_fdiv))

# all events
  # Emergence_of_Canidae
historical_trait_space_funspace_Emergence_of_Canidae <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Emergence_of_Canidae`, n_divisions = 300)
plot(x = historical_trait_space_funspace_Emergence_of_Canidae, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_Emergence_of_Canidae)
Emergence_of_Canidae_fric <- historical_trait_space_funspace_Emergence_of_Canidae$FD$groups$FRich["retained"]
Emergence_of_Canidae_fdiv <- historical_trait_space_funspace_Emergence_of_Canidae$FD$groups$FDiv["retained"]
  
  # Decline_of_broadleaf_forest
historical_trait_space_funspace_Decline_of_broadleaf_forest <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Decline_of_broadleaf_forest`, n_divisions = 300)
plot(x = historical_trait_space_funspace_Decline_of_broadleaf_forest, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_Decline_of_broadleaf_forest)
Decline_of_broadleaf_forest_fric <- historical_trait_space_funspace_Decline_of_broadleaf_forest$FD$groups$FRich["retained"]
Decline_of_broadleaf_forest_fdiv <- historical_trait_space_funspace_Decline_of_broadleaf_forest$FD$groups$FDiv["retained"]

  # Expansion_of_sparse_temperate_forest
historical_trait_space_funspace_Expansion_of_sparse_temperate_forest <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Expansion_of_sparse_temperate_forest`, n_divisions = 300)
plot(x = historical_trait_space_funspace_Expansion_of_sparse_temperate_forest, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_Expansion_of_sparse_temperate_forest)
Expansion_of_sparse_temperate_forest_fric <- historical_trait_space_funspace_Expansion_of_sparse_temperate_forest$FD$groups$FRich["retained"]
Expansion_of_sparse_temperate_forest_fdiv <- historical_trait_space_funspace_Expansion_of_sparse_temperate_forest$FD$groups$FDiv["retained"]
  
  # Catgap
historical_trait_space_funspace_Catgap <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Catgap`, n_divisions = 300)
plot(x = historical_trait_space_funspace_Catgap, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_Catgap)
Catgap_fric <- historical_trait_space_funspace_Catgap$FD$groups$FRich["retained"]
Catgap_fdiv <- historical_trait_space_funspace_Catgap$FD$groups$FDiv["retained"]

  # Catgap_filled
historical_trait_space_funspace_Catgap_filled <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Catgap_filled`, n_divisions = 300)
plot(x = historical_trait_space_funspace_Catgap_filled, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_Catgap_filled)
Catgap_filled_fric <- historical_trait_space_funspace_Catgap_filled$FD$groups$FRich["retained"]
Catgap_filled_fdiv <- historical_trait_space_funspace_Catgap_filled$FD$groups$FDiv["retained"]
  
  # Mid_miocene_climatic_optimum
historical_trait_space_funspace_Mid_miocene_climatic_optimum <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Mid_miocene_climatic_optimum`, n_divisions = 300)
plot(x = historical_trait_space_funspace_Mid_miocene_climatic_optimum, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_Mid_miocene_climatic_optimum)
Mid_miocene_climatic_optimum_fric <- historical_trait_space_funspace_Mid_miocene_climatic_optimum$FD$groups$FRich["retained"]
Mid_miocene_climatic_optimum_fdiv <- historical_trait_space_funspace_Mid_miocene_climatic_optimum$FD$groups$FDiv["retained"]

  # Extinction_of_Hesperocyoninaes
historical_trait_space_funspace_Extinction_of_Hesperocyoninaes <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Extinction_of_Hesperocyoninaes`, n_divisions = 300)
plot(x = historical_trait_space_funspace_Extinction_of_Hesperocyoninaes, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_Extinction_of_Hesperocyoninaes)
Extinction_of_Hesperocyoninaes_fric <- historical_trait_space_funspace_Extinction_of_Hesperocyoninaes$FD$groups$FRich["retained"]
Extinction_of_Hesperocyoninaes_fdiv <- historical_trait_space_funspace_Extinction_of_Hesperocyoninaes$FD$groups$FDiv["retained"]
  
  # C4_grassland_expansion
historical_trait_space_funspace_C4_grassland_expansion <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`C4_grassland_expansion`, n_divisions = 300)
plot(x = historical_trait_space_funspace_C4_grassland_expansion, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_C4_grassland_expansion)
C4_grassland_expansion_fric <- historical_trait_space_funspace_C4_grassland_expansion$FD$groups$FRich["retained"]
C4_grassland_expansion_fdiv <- historical_trait_space_funspace_C4_grassland_expansion$FD$groups$FDiv["retained"]

  # Maximum_diversity_of_horses
historical_trait_space_funspace_Maximum_diversity_of_horses <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Maximum_diversity_of_horses`, n_divisions = 300)
plot(x = historical_trait_space_funspace_Maximum_diversity_of_horses, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_Maximum_diversity_of_horses)
Maximum_diversity_of_horses_fric <- historical_trait_space_funspace_Maximum_diversity_of_horses$FD$groups$FRich["retained"]
Maximum_diversity_of_horses_fdiv <- historical_trait_space_funspace_Maximum_diversity_of_horses$FD$groups$FDiv["retained"]
  
  # Isthmus_of_Panama
historical_trait_space_funspace_Isthmus_of_Panama <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Isthmus_of_Panama`, n_divisions = 300)
plot(x = historical_trait_space_funspace_Isthmus_of_Panama, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_Isthmus_of_Panama)
Isthmus_of_Panama_fric <- historical_trait_space_funspace_Isthmus_of_Panama$FD$groups$FRich["retained"]
Isthmus_of_Panama_fdiv <- historical_trait_space_funspace_Isthmus_of_Panama$FD$groups$FDiv["retained"]

  # Extinction_of_Borophaginaes
historical_trait_space_funspace_Extinction_of_Borophaginaes <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Extinction_of_Borophaginaes`, n_divisions = 300)
plot(x = historical_trait_space_funspace_Extinction_of_Borophaginaes, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_Extinction_of_Borophaginaes)
Extinction_of_Borophaginaes_fric <- historical_trait_space_funspace_Extinction_of_Borophaginaes$FD$groups$FRich["retained"]
Extinction_of_Borophaginaes_fdiv <- historical_trait_space_funspace_Extinction_of_Borophaginaes$FD$groups$FDiv["retained"]
  
  # Megafauna_extinction
historical_trait_space_funspace_Megafauna_extinction <-  funspace(x = historical_funspace_df_num, group.vec = historical_funspace_df$`Megafauna_extinction`, n_divisions = 300)
plot(x = historical_trait_space_funspace_Megafauna_extinction, type = "groups", which.group = "retained",quant.plot = TRUE, globalContour = T, pnt = T, pnt.cex = 0.1, pnt.col = rgb(0.2,0.8,0.1,alpha=0.2), axis.title.line = 3, axis.title.x = "RBL", axis.title.y = "Mass")
summary(historical_trait_space_funspace_Megafauna_extinction)
Megafauna_extinction_fric <- historical_trait_space_funspace_Megafauna_extinction$FD$groups$FRich["retained"]
Megafauna_extinction_fdiv <- historical_trait_space_funspace_Megafauna_extinction$FD$groups$FDiv["retained"]

# Collate all results
historical_results <- tibble::tibble(
  Scenario = c("Global", "Emergence of Canidae", "Decline of broadleaf forest", "Expansion of sparse temperate forest", "Catgap", "Catgap", "Mid miocene climatic optimum",
               "Extinction of Hesperocyoninaes", "C4 grassland expansion", "Maximum diversity of horses", "Isthmus of Panama", "Extinction of Borophaginaes", "Megafauna extinction"),
  FRic = c(historical_fric, Emergence_of_Canidae_fric, Decline_of_broadleaf_forest_fric, Expansion_of_sparse_temperate_forest_fric, Catgap_fric, Catgap_filled_fric,
           Mid_miocene_climatic_optimum_fric, Extinction_of_Hesperocyoninaes_fric, C4_grassland_expansion_fric, Maximum_diversity_of_horses_fric, Isthmus_of_Panama_fric,
           Extinction_of_Borophaginaes_fric, Megafauna_extinction_fric),
  FDiv = c(historical_fdiv, Emergence_of_Canidae_fdiv, Decline_of_broadleaf_forest_fdiv, Expansion_of_sparse_temperate_forest_fdiv, Catgap_fdiv, Catgap_filled_fdiv,
           Mid_miocene_climatic_optimum_fdiv, Extinction_of_Hesperocyoninaes_fdiv, C4_grassland_expansion_fdiv, Maximum_diversity_of_horses_fdiv, Isthmus_of_Panama_fdiv,
           Extinction_of_Borophaginaes_fdiv, Megafauna_extinction_fdiv))

historical_results
```

## Plot results
```{r plot historical}
# All periods
events_list <- list(
  Emergence_of_Canidae = historical_trait_space_funspace_Emergence_of_Canidae,
  Decline_of_broadleaf_forest = historical_trait_space_funspace_Decline_of_broadleaf_forest,
  Expansion_of_sparse_temperate_forest = historical_trait_space_funspace_Expansion_of_sparse_temperate_forest,
  Catgap = historical_trait_space_funspace_Catgap,
  Catgap_filled = historical_trait_space_funspace_Catgap_filled,
  Mid_miocene_climatic_optimum = historical_trait_space_funspace_Mid_miocene_climatic_optimum,
  Extinction_of_Hesperocyoninaes = historical_trait_space_funspace_Extinction_of_Hesperocyoninaes,
  C4_grassland_expansion = historical_trait_space_funspace_C4_grassland_expansion,
  Maximum_diversity_of_horses = historical_trait_space_funspace_Maximum_diversity_of_horses,
  Isthmus_of_Panama = historical_trait_space_funspace_Isthmus_of_Panama,
  Extinction_of_Borophaginaes = historical_trait_space_funspace_Extinction_of_Borophaginaes,
  Megafauna_extinction = historical_trait_space_funspace_Megafauna_extinction
)

par(mfrow=c(3,4))

for (nm in names(events_list)) {

  plot(
    x = events_list[[nm]],
    type = "groups",
    which.group = "retained",
    quant.plot = TRUE,
    globalContour = TRUE,
    pnt = TRUE,
    pnt.cex = 0.1,
    pnt.col = rgb(0.2,0.8,0.1,alpha=0.2),
    axis.title.x = "RBL",
    axis.title.y = "Mass",
    axis.title.cex = 0.8,
    axis.cex = 0.5
  )

  title(gsub("_"," ",nm), cex.main = 1.1)
}

# Only major extinction events
extinction_events_list <- list(
  Extinction_of_Hesperocyoninaes = historical_trait_space_funspace_Extinction_of_Hesperocyoninaes,
  Extinction_of_Borophaginaes = historical_trait_space_funspace_Extinction_of_Borophaginaes,
  Megafauna_extinction = historical_trait_space_funspace_Megafauna_extinction
)

par(mfrow=c(1,3))

for (nm in names(extinction_events_list)) {

  plot(
    x = events_list[[nm]],
    type = "groups",
    which.group = "retained",
    quant.plot = TRUE,
    globalContour = TRUE,
    pnt = TRUE,
    pnt.cex = 0.1,
    pnt.col = rgb(0.2,0.8,0.1,alpha=0.2),
    axis.title.x = "RBL",
    axis.title.y = "Mass",
    axis.title.cex = 0.7,
    axis.cex = 0.7
  )

  title(gsub("_"," ",nm), cex.main = 1)
}

# Borophaginae decline
Borophaginae_events_list <- list(
  Maximum_diversity_of_horses = historical_trait_space_funspace_Maximum_diversity_of_horses,
  Isthmus_of_Panama = historical_trait_space_funspace_Isthmus_of_Panama,
  Extinction_of_Borophaginaes = historical_trait_space_funspace_Extinction_of_Borophaginaes
)

par(mfrow=c(1,3))

for (nm in names(Borophaginae_events_list)) {

  plot(
    x = events_list[[nm]],
    type = "groups",
    which.group = "retained",
    quant.plot = TRUE,
    globalContour = TRUE,
    pnt = TRUE,
    pnt.cex = 0.1,
    pnt.col = rgb(0.2,0.8,0.1,alpha=0.2),
    axis.title.x = "RBL",
    axis.title.y = "Mass",
    axis.title.cex = 0.7,
    axis.cex = 0.7
  )

  title(gsub("_"," ",nm), cex.main = 1)
}
```